Locality and Effect Reflection

### 2025-07-01 paper.tex
Write more

\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{fontawesome5}
\usepackage{mathpartir}
\usepackage{sourcecodepro}

@@

\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catop}{\cat^{\mathtt{op}}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{#1 \Rightarrow #2}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflect}[2]{\reflectname_{#1}(#2)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reify}[2]{\reifyname_{#1}(#2)}
\newcommand{\lock}{\text{\faLock}}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}

@@

\begin{abstract}
  Algebraic effects and handlers allow the programmer to extend their
  languages \emph{ambient monad} \textbf{--} the monad describing which
  effects ordinary functions can perform \textbf{--} with additional
  effectful operations.

@@

  We introduce an alternative presentation of algebraic effects and
  handlers, \emph{effect reflection}, that provides, for any algebraic
  theory $\Sigma$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}$ is an inductive type representing the free
  algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
  to inject terms of $\Sigma$ into the current ambient
  monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
  $\Sigma$. This presentation falls out naturally from a simple
  denotational semantics for algebraic effects in terms of presheaves
  over bounded Lawvere theories.

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing monadic code to be written in direct
  style. This provides all the benefits of typed algebraic effects and
  handlers without requiring an effect system. We demonstrate effect
  reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
  examples of programming with effects.
\end{abstract}

@@

let handle (f : unit -> 'a) : 'a =
  match f () with
  | x -> fun _ -> x
  | effect Get, k -> fun s -> continue k s s
  | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=ocaml]{f ()} function call
the ambient monad is extended with two additional effectful operations:
\begin{lstlisting}[style=ocaml]
let get () : int = perform Get
let set (i : int) : unit = perform (Set i)
\end{lstlisting}

@@

@@

let handle (f : Handler(State).t @ local -> 'a) : 'a =
  let rec loop s : 'a Term(State).t -> 'a  = function
    | Return x -> x
    | Op(Get, k) -> loop s (k s)
    | Op(Set s', k) -> loop s' (k ())
  in
  loop (State_eff.reify f)
\end{lstlisting}
which calls \lstinline[style=ocaml]{f} and passes it a local value of
type \lstinline[style=ocaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=ocaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=ocaml]
let get (h : Handler(State).t @ local) : int =
  State_eff.reflect (Op(Get, fun x -> x)) h

let set (h : Handler(State).t @ local) (i : int) : unit =
  State_eff.reflect (Op(Set i, fun x -> x)) h
\end{lstlisting}

@@

\begin{itemize}
\item The global modality $\glob$ on such presheaves is defined such
  that $\glob(P)(L) = P(\initial)$. In other words, for a given presheaf
  $P$ it gives the constant presheaf that always ignores the ambient
  monad and instead returns $P$ at $\initial$, where $\initial$ is the
  algebraic theory with no operations or equations. This basically
  restricts a type to its terms that do not depend on the ambient monad.

\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebra. We use
  $\mon$ to define the semantics of computations.

@@

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=ocaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=ocaml]
let leak_local x =
  let (o @ local) = Some x in
  o
(*@ \textcolor{red}{Error: This value is local and would escape its region.} @*)
\end{lstlisting}

@@

By default, modes are deep: a local \lstinline[style=ocaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=ocaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=ocaml]
type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=ocaml]{global}
modality. The contents of the \lstinline[style=ocaml]{g} field is always
global, for example:
\begin{lstlisting}[style=ocaml]
let return_global x =
  let (r @ local) = { g = x } in
  r.g
\end{lstlisting}
is allowed and gives a function of type:
\begin{lstlisting}[style=ocaml]
val return_global : 'a @@ global -> 'a @@ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=ocaml]{global} modality,
there is no \lstinline[style=ocaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region is will
implicit cause the second one to leave that region as well. This means
the \lstinline[style=ocaml]{global} mode is always deep.

@@

\begin{lstlisting}[style=ocaml]
module type Op = sig type 'a t end

module Term (O : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler(O : Op) : sig
  type 'a t
end

module Reflection (O : Op) : sig

  val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

  val reflect : 'a Term(Op).t -> (Handler(O).t @ local -> 'a)

@@

  val reflect_local : 'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local

\begin{lstlisting}[style=ocaml]
let perform (op : 'a O.t) : 'a =
  reflect (Op(op, fun x -> x))
\end{lstlisting}

@@

@@

\begin{lstlisting}[style=ocaml]
module Sum (L : Op) (R : Op) : sig
  type 'a t =
    | Left : 'a L.t -> 'a t
    | Right : 'a R.t -> 'a t
end

@@

As a Lawvere theory, \lstinline[style=ocaml]{Sum(L)(R)} is the sum of
\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}. This means
that the terms of its free algebra $T_{L + R}(\tau)$ are equivalent to
values of the initial algebra $\mu X. T_L(\tau + F_L(X)$. We can expose
this fact via the following functions:
\begin{lstlisting}
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t

  let rec reflect_left : 'a Term(Sum(L)(R)).t -> (Handler(L).t @ local -> 'a Half_term(L)(R).t)
   = function
    | Return v -> fun _ -> Return v
    | Op(Left op, k) -> fun l -> L.reflect (Op(op, (fun res -> reflect_left (k res) l)))
    | Op(Right op, k) -> fun _ -> Op(op, (fun res -> reflect_left (k res)))

  let rec reify_left : (Handler(L).t @ local -> 'a Half_term(L)(R).t) -> 'a Term(Sum(L)(R)).t =
    fun h ->
      let rec loop = function
        | Return (Return x) -> Return x
        | Return (Op(r, k)) -> Op(Right r, fun res -> reify_left (k res))
        | Op(l, k) -> Op(Left l, fun res -> loop (k res))
      in
      loop (L.reify h)

\subsection{Examples}

\subsubsection{State}

\subsubsection{Concurrency}

\subsubsection{Generators and concurrency}

@@

>>>>

\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{fontawesome5}
\usepackage{mathpartir}
\usepackage{sourcecodepro}
\usepackage{amsthm}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

@@

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{#1 \Rightarrow #2}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[4]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducesto}{\rightsquigarrow}

@@

\begin{abstract}
  Algebraic effects and handlers allow the programmer to extend their
  language's \emph{ambient monad} \textbf{--} the monad describing which
  effects ordinary functions can perform \textbf{--} with additional
  effectful operations.

@@

  We introduce an alternative presentation of algebraic effects and
  handlers, \emph{effect reflection}, that provides, for any algebraic
  signature $\Sigma$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}$ is an inductive type representing the free
  algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
  to inject terms of $\Sigma$ into the current ambient
  monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
  $\Sigma$. This presentation falls out naturally from a simple
  denotational semantics for algebraic effects in terms of presheaves
  over bounded Lawvere theories.

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing effectful code to be written in
  direct style. This provides all the benefits of typed algebraic
  effects and handlers without requiring an effect system. We
  demonstrate effect reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
  examples of programming with effects.
\end{abstract}

@@

let handle_state (f : unit -> 'a) : 'a =
  match f () with
  | x -> fun _ -> x
  | effect Get, k -> fun s -> continue k s s
  | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=ocaml]{f ()} function call
the ambient monad is extended with two additional effectful operations:
\begin{lstlisting}[style=ocaml]
let get () : int = perform Get
let set (i : int) : unit = perform (Set i)
\end{lstlisting}

@@

\todo{Restrict the above claim to signatures}

@@

let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
  let rec loop s : 'a Term(State).t @ local -> 'a  = function
    | Return x -> x
    | Op(Get, k) -> loop s (k s)
    | Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
which calls \lstinline[style=ocaml]{f} and passes it a local value of
type \lstinline[style=ocaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=ocaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=ocaml]
let get (h : Handler(State).t @ local) : int =
  State_eff.perform Get h

let set (h : Handler(State).t @ local) (i : int) : unit =
  State_eff.perform (Set i) h
\end{lstlisting}

@@

\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebra. We use
  $\mon$ to define the semantics of computations.

\item The global modality $\glob$ on such presheaves is defined such
  that $\glob(P)(L) = P(\initial)$. In other words, for a given presheaf
  $P$ it gives the constant presheaf that always ignores the ambient
  monad and instead returns $P$ at $\initial$, where $\initial$ is the
  algebraic theory with no operations or equations. This allows us to
  restrict a type to its terms that do not depend on the ambient monad.

@@

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=ocaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=ocaml]
let leak_local x =
  let (o @ local) = Some x in
  o
(*@ \textcolor{red}{Error: o} @*)
       (*@ \textcolor{red}{\textasciicircum} @*)
(*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

@@

By default, modes are deep: a local \lstinline[style=ocaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=ocaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=ocaml]
type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=ocaml]{global}
modality. The contents of the \lstinline[style=ocaml]{g} field is always
global, for example:
\begin{lstlisting}[style=ocaml]
let project_global (r @ local) =
  r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=ocaml]
val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=ocaml]{global} modality,
there is no \lstinline[style=ocaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region it will
implicitly cause the second one to leave that region as well. This means
the \lstinline[style=ocaml]{global} mode is always deep.

@@

The types and operations of our effect reflection API are parameterized
by type constructors representing the operations of an algebraic
effect. In OxCaml, such higher-kinded interfaces must be built using
functors, so we start with a simple module type
\lstinline[style=ocaml]{Op} to act as the parameter type for these
functors:
\begin{lstlisting}[style=ocaml]
module type Op = sig type 'a t end
\end{lstlisting}
The fundamental types of the API are the \lstinline[style=ocaml]{Term}
and \lstinline[style=ocaml]{Handler} types.
\begin{lstlisting}[style=ocaml]
module Term (O : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler(O : Op) : sig
  type t
end
\end{lstlisting}
For an effect signature \lstinline[style=ocaml]{O},
\lstinline[style=ocaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{??}. \lstinline[style=ocaml]{Handler(O).t} is the type of
mappings from the operations of \lstinline[style=ocaml]{O} into the
operations of the current ambient monad. Alternatively,
\lstinline[style=ocaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=ocaml]{'a Term(O).t}} \qquad
  \cong \qquad \text{\lstinline[style=ocaml]{Handler(O).t @ local -> 'a}}
\end{equation*}
Due to the absence of mode polymorphism in OxCaml, we expose this
isomorphism as three functions: \lstinline[style=ocaml]{reify} and
\lstinline[style=ocaml]{reify_local} which implement the left-to-right
direction at the \lstinline[style=ocaml]{global} and
\lstinline[style=ocaml]{local} modes respectively; and
\lstinline[style=ocaml]{perform}, which is equivalent to the
right-to-left direction but makes for a more convenient primitive in
practice. \lstinline[style=ocaml]{perform} amounts to reflection of
individual operations as opposed to full terms.
\begin{lstlisting}[style=ocaml]
module Reflection (O : Op) : sig
  val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

@@

  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
end
\end{lstlisting}
Using \lstinline[style=ocaml]{perform} we can implement reflection of
terms as:
\begin{lstlisting}[style=ocaml]
let rec reflect t h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect (k (perform op h)) h

let rec reflect_local (t @ local) h =
  match t with
  | Return x -> x
  | Op(op, k) -> reflect_local (k (perform op h)) h
\end{lstlisting}
with types:
\begin{lstlisting}[style=ocaml]
val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)

val reflect_local :
   'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local
\end{lstlisting}

\subsection{Example: State}

As shown in the introduction we can implement an effect handler for
state using this API. We start by defining the operations of the effect:
\begin{lstlisting}[style=ocaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end
\end{lstlisting}
Then we instantiate the reflection isomorphism for that effect and use
it to write a handler:
\begin{lstlisting}[style=ocaml]
module State_eff = Reflection(State)

let handle_state f i =
  let rec loop s = function
    | Term(State).Return x -> x
    | Term(State).Op(Get, k) -> loop s (k s)
    | Term(State).Op(Set s', k) -> loop s' (k ())
  in
  loop i (State_eff.reify_local f)
\end{lstlisting}
Finally we add an alias for the handler type and some convenient
wrappers for the generic effects:
\begin{lstlisting}[style=ocaml]
type handler = Handler(State).t

let get h =
  State_eff.perform Get h

let set h i =
  State_eff.perform (Set i) h
\end{lstlisting}
Giving us the following interface:
\begin{lstlisting}[style=ocaml]
type handler

val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
val get : handler @ local -> int
val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=ocaml]{reify_local} to allow
the function passed to \lstinline[style=ocaml]{handle_state} to be local.

This shows how to implement state using effect reflection, but note that
the same interface can also be implemented direcly using mutable state
and locality:
\begin{lstlisting}[style=ocaml]
type handler = int ref

let handle_state f i = f (ref i)
let get r = !r
let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state monad. In
some sense, the improved reasoning ability of the state monad over
ordinary mutable state is precisely captured by the notion of locality.

This transformation from effect relection to local mutable state works
for any \emph{tail-resumptive} effect handler.

\subsection{Lightweight effect polymorphism}

The following function uses state to calculate the total of a list of
integers:
\begin{lstlisting}[style=ocaml]
let total l =
  handle_state (fun h ->
    List.iter (fun x -> set h (get h + x)) l;
    get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=ocaml]{List.iter}:
\begin{lstlisting}[style=ocaml]
val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=ocaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=ocaml]{local} handler \lstinline[style=ocaml]{h} and
use it to perform some effects.

This ability for higher-order functions like
\lstinline[style=ocaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by ??
et al.\cite{??} without requiring a separate notion of second-class
function.

\subsection{Example: Generators}

A more realistic use of effect reflection is a \emph{generator}
effect. This effect has a single operation:
\begin{lstlisting}[style=ocaml]
module Gen = struct
  type 'a t =
    | Gen : int -> unit t
end
\end{lstlisting}
Along with a handler that turns computations using
\lstinline[style=ocaml]{Gen} into streams of integers.
\begin{lstlisting}[style=ocaml]
module Gen_eff = Reflection(Gen)

type ints =
  | Finished
  | More of int * (unit -> ints)

let handle_gen f =
  let rec loop = function
    | Term(Gen).Return () -> Finished
    | Term(Gen).Op(Gen i, k) -> More(i, k)
  in
  loop (Gen_eff.reify f)
\end{lstlisting}

Unlike the state handler, here we must use
\lstinline[style=ocaml]{reify} rather than
\lstinline[style=ocaml]{reify_local} because our intention is that the
resulting stream of integers be a global value that can escape the
current region. Since it closes over the continuation
\lstinline[style=ocaml]{k}, we must use \lstinline[style=ocaml]{reify}
and require that \lstinline[style=ocaml]{f} be global:
\begin{lstlisting}[style=ocaml]
val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
\end{lstlisting}

\subsection{Example: Asynchronous I/O}

Perhaps the most common use case for algebraic effects is for
concurrency and asynchronous I/O. Implementing asynchronous I/O from
scratch is beyond the scope of this paper, but we can build a simple
version on top of an existing monadic concurrency library. Assuming we
have a monadic future type \lstinline[style=ocaml]{'a Deferred.t} we can
write concurrent code in direct style by using an effect with a single
\lstinline[style=ocaml]{Await} operation that waits for a future to be
completed:
\begin{lstlisting}[style=ocaml]
module Await : sig
  type 'a t =
    | Await : 'a Deferred.t -> 'a t
end
\end{lstlisting}

\begin{lstlisting}[style=ocaml]
module Await_eff = Reflection(Await)

let handle_await f =
  let rec loop = function
    | Term(Await).Return x -> Deferred.return x
    | Term(Await).Op(Await d, k) ->
        Deferred.bind d (fun x -> loop (k x))
  in
  loop (Await_eff.reify f)

let await h d =
  Await_eff.perform (Await d) h
\end{lstlisting}

Using this we can write direct-style code using asynchronous I/O:
\begin{lstlisting}[style=ocaml]
let copy_file h src dst =
  let s = await h (File.read src) in
  await h (File.write dst s)
\end{lstlisting}

@@

\subsection{Nested effect handlers}

A key feature of algebraic effects is how easy they are to compose. We
can use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=ocaml]
handle_await (fun ch ->
  handle_state i (fun sh ->
    List.iter files (fun fl ->
      let s = await ch (File.read fl) in
      let i = int_of_string s in
      set sh (i + get sh));
    get sh))
\end{lstlisting}

This relies on the fact that the \lstinline[style=ocaml]{handle_state}
uses \lstinline[style=ocaml]{reify_local} and so is able to operate on a
local computation that closes over the local handler value
\lstinline[style=ocaml]{ch}.

If we tried to do something similar using our
\lstinline[style=ocaml]{handle_gen} handler, then we get and error:
\begin{lstlisting}[style=ocaml]
handle_await (fun ch ->
  handle_gen i (fun gh ->
    List.iter files (fun fl ->
      let s = await ch (File.read fl) in
      let i = int_of_string s in
      gen gh i)))
(*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                     (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
(*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

We could change the definition of \lstinline[style=ocaml]{handle_gen} to
use \lstinline[style=ocaml]{reify_local} instead of
\lstinline[style=ocaml]{reify}, but then we would lose the property that
the generator we produce can be passed around freely. Instead we need to
build a handler that can forward the \lstinline[style=ocaml]{Await}
operation to different effect handlers at different points in the
program's execution.

@@

Composing handlers, like \lstinline[style=ocaml]{handle_gen}, that use
\lstinline[style=ocaml]{reify} rather than
\lstinline[style=ocaml]{reify_local} is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=ocaml]{Await} and \lstinline[style=ocaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=ocaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=ocaml]{Await} operations. Whenever we force part of the
stream we'll need to provide a handler for those
\lstinline[style=ocaml]{Await} operations.

To acheive this more general form of composition, we require an
additional capability in our effect reflection API. We need to be able
to manipulate \lstinline[style=ocaml]{Handler} types for sums of
effects.
\begin{lstlisting}[style=ocaml]
module Sum (L : Op) (R : Op) : sig
  type 'a t =
    | Left : 'a L.t -> 'a t
    | Right : 'a R.t -> 'a t
end

@@

\lstinline[style=ocaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}.

Terms in the free algebra of the sum of two effects
(i.e. \lstinline[style=ocaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{??}. We can definet this type as:
\begin{lstlisting}[style=ocaml]
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
end
\end{lstlisting}
and expose the isomorphism between
\lstinline[style=ocaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
and \lstinline[style=ocaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
\begin{lstlisting}[style=ocaml]
module Half_reflection (L : Op) (R : Op) : sig
  module L_eff = Reflection(L)
  module R_eff = Reflection(R)
  module Sum_eff = Reflection(Sum(L)(R))
  module Proj = Project(L)(R)

  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
    let rec loop lh = function
      | Term(Sum(L)(R)).Return ->
          Half_term(L)(R).Return v
      | Term(Sum(L)(R)).Op(Left op, k) ->
          loop lh (k (L_eff.perform op lh))
      | Term(Sum(L)(R)).Op(Right op, k) ->
          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
    in
    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh

  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
    match f lh with
    | Return x -> x
    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
end
\end{lstlisting}

Now we can use this machinery to conveniently build a handler of
\lstinline[style=ocaml]{Gen} that also forwards operations from
\lstinline[style=ocaml]{Await}:
\begin{lstlisting}[style=ocaml]
module Await_gen_eff = Half_reflection(Await)(Gen)

type aints =
  | Finished
  | More of int * (Handler(Await).t @ local -> ints)

let handle_gen_in_await f ah =
    let rec loop = function
    | Half_term(Await)(Gen).Return () -> Finished
    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, k ())
  in
  loop (Await_gen_eff.reify f)
\end{lstlisting}
where \lstinline[style=ocaml]{handle_gen_in_await} has the type:
\begin{lstlisting}[style=ocaml]
val handle_gen_in_await :
  (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
  -> Handler(Await).t @ local
  -> aints
\end{lstlisting}

The literature on lexical effect handlers often ignores this issue. For
example, \cite{??} proposed a system which only supported directly
nesting effect handlers and had no mechanism for writing a composable
handler for generators that could be moved between different underlying
asynchronous I/O handlers.

This is because this composition requiring forwarding is much less
common than nesting and it fundamentally requires each use of the
\lstinline[style=ocaml]{Await} operation to perform work proportional to
the number of handlers it is forwarded through -- often referred to as a
``linear search''. There has been a tendency to treat this linear search
as some form of deficiency of algebraic effect handlers, but it is
simply a classic trade-off between the extra expressivity allowed by an
indirection and the runtime cost of traversing that indirection. With
our effect reflection API we are able to use nesting for the cases that
it supports -- avoiding the cost of the indirection -- and fall back to
forwarding for the more general cases.

\subsection{Parameterized effects and handlers}

The state handlers in this paper so far have all been restricted to
\lstinline[style=ocaml]{int} state. This is a limitation of the effect
reflection API that we have been using. We would like to support
\emph{parameterized effects} -- allowing a single handler to be used
with a whole family of effects. For example, a state handler
parameterized by the type of the state.

We can support parameterized effects by adding a module type for them,
along with parameterized versions of \lstinline[style=ocaml]{Term} and
\lstinline[style=ocaml]{Handler}:
\begin{lstlisting}[style=ocaml]
module type Op1 = sig type ('a, 'e) t end

module Term1 (O : Op1) : sig
  type ('a, 'e) t =
    | Return : 'a @@ global -> 'a t
    | Op : ('r, 'e) O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler1(O : Op1) : sig
  type 'e t
end
\end{lstlisting}
with a corresponding parameterized version of
\lstinline[style=ocaml]{Reflection}. We can then define a parameterized
state effect:
\begin{lstlisting}[style=ocaml]
module State = struct
  type ('a, 'b) t =
    | Get : 'b t
    | Set : 'b -> unit t
end
\end{lstlisting}
and write a handler that provides a parameterized interface for the
state effect:
\begin{lstlisting}[style=ocaml]
type 's handler

val handle : ('s handler @ local -> 'a) @ local -> 's -> 'a
val get : 's handler @ local -> 's
val set : 's handler @ local -> 's -> unit
\end{lstlisting}

The need for a new copy of the interface for each type constructor arity
is an artifact of OCaml's approach to higher-kinded types. Languages
with direct support for higher-kinded types should be able to support
parameterized effects of arbitrary arity with a single interface.

@@

\section{Appendix}
